---
title: "Final project"
author: "Reese Quillian"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project Topic: Crime, Cybersecurity

```{r}
library(tidyverse)
setwd("C:/Users/Student/OneDrive - University of Virginia/Desktop/SYS2202/Final")

# complaints dataset
complaints <- read_csv("NYPD_Complaint_Data_Current__Year_To_Date_.csv")

# income dataset
income <- read_csv("Citywide_Payroll_Data__Fiscal_Year_.csv")
```

Note: All steps must be demonstrated and implemented in R. 

# Describe what the datasets are about and list one main questions/hypothesis you want to test with this data.

Question: How does the level of wealth in an area affect the frequency of crimes committed?
Sub-Question: What about the types of crimes being committed?

# Relevant variables

## What are the types of those variables? How do you determine that?

Complaints dataset: LAW_CAT_CD, BORO_NM

```{r}
complaints %>%
  select(LAW_CAT_CD, BORO_NM) %>%
  glimpse()
```

Income dataset: Work Location Borough, Base Salary
```{r}
income %>%
  select(`Work Location Borough`, `Base Salary`) %>%
  glimpse()
```


# Search for evidence by visualising, transforming, and modeling your data

## What type of variation occurs within each variable?

### Variables to examine: LAW_CAT_CD, BORO_NM

#### Visualize distributions (Barcharts, Histograms)

We will start by just looking at crimes by borough before factoring in income.

```{r}
complaints %>%
  select(CMPLNT_FR_DT, KY_CD, LAW_CAT_CD, BORO_NM) %>%
  filter(!is.na(BORO_NM)) %>%
  group_by(BORO_NM) %>%
  summarize(count = n()) %>%
  ggplot(aes(x=reorder(BORO_NM, -(count)), y=count, fill=BORO_NM)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  theme(legend.position = "none") + 
  ggtitle("Number of Crime Complaints by Borough") + xlab("Borough Name")
```


It can be observed that the Bronx experiences the highest number of complaints in this dataset. Now we can look into the number of complaints by the level of offense (felony, misdemeanor, violation). First we will look at the distribution of LAW_CAT_CD using a bar graph.

```{r}
complaints %>%
  select(CMPLNT_FR_DT, KY_CD, LAW_CAT_CD, BORO_NM) %>%
  filter(!is.na(LAW_CAT_CD)) %>%
  group_by(LAW_CAT_CD) %>%
  summarize(count = n()) %>%
  ggplot(aes(x=reorder(LAW_CAT_CD, -(count)), y=count, fill=LAW_CAT_CD)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  theme(legend.position = "none") + 
  ggtitle("Number of Crime Complaints by Type") + xlab("Type")
```


```{r}
complaints %>%
  select(CMPLNT_FR_DT, KY_CD, LAW_CAT_CD, BORO_NM) %>%
  filter(!is.na(BORO_NM)) %>%
  group_by(BORO_NM, LAW_CAT_CD) %>%
  summarize(count = n()) %>%
  
  ggplot(aes(x=reorder(BORO_NM, -(count)), y=count, fill=LAW_CAT_CD)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  ggtitle("Number of Crime Complaints by Borough \n and Level of Offense") +
  xlab("Borough Name")
```

```{r}
complaints %>%
  select(CMPLNT_FR_DT, KY_CD, LAW_CAT_CD, BORO_NM) %>%
  filter(!is.na(BORO_NM)) %>%
  group_by(BORO_NM, LAW_CAT_CD) %>%
  summarize(count = n()) %>%
  
  ggplot(aes(x=reorder(BORO_NM, -(count)), y=count, fill=LAW_CAT_CD)) + 
  geom_bar(position = "fill", stat="identity") + 
  coord_flip() +
  ggtitle("Number of Crime Complaints by Borough \n and Level of Offense") +
  xlab("Borough Name") + ylab("Percent")
```

#### Check Unusual values


#### Identify and Handle Missing values

```{r}
complaints %>%
  filter(is.na(BORO_NM)) %>%
  count()
```
There are 1151 observations that are missing the name of the borough. This is only (approximately) 0.2% of all total complaints, so they will be removed. 

  
### Variables to examine: Work Location Borough, Base Salary

#### Visualize distributions (Barcharts, Histograms)

Base Salary is listed in multiple different ways (yearly, monthly, hourly).
We will eventually need to convert all observations into a yearly salary.

Before performing the conversion, we should look at how the data is distributed between each type, and the distribution of salaries within each type. 

```{r}
income_annual <- income %>%
  # rename columns so they are easier to work with
  rename(boro = `Work Location Borough`, 
         salary = `Base Salary`, 
         basis = `Pay Basis`) %>%
  # we only want to look at individuals who are actively working, otherwise the   data will be skewed
  filter(`Leave Status as of June 30` == "ACTIVE") %>%
  select(boro, basis, salary) %>%
  filter(basis == 'per Annum')
income_annual
```

```{r}
# Distribution of annual salaries
income_annual %>%
  filter(basis == 'per Annum') %>%
  ggplot(aes(x=salary)) + 
  geom_histogram() + ggtitle("Distribution of Annual Salaries")
```

The histogram is right skewed, meaning there are some high outliers. These values may need to be removed/assigned a new value later in the analysis.

```{r}
summarize(income_annual, median(salary), mean(salary), IQR(salary), sd(salary))
summary(income_annual$salary)
```

```{r}
# Determining outliers
income_annual %>%
  filter(salary > quantile(salary,0.75) + 1.5*IQR(salary))
  
```


```{r}
# Cleaning borough variable
library(stringr)
income_annual %>%
  mutate(BORO_NM = str_to_upper(boro)) %>%
  filter(BORO_NM == 'BRONX' | BORO_NM == 'BROOKLYN' | BORO_NM == 'QUEENS' |
         BORO_NM == 'MANHATTAN') %>%
  filter(basis == 'per Annum') %>%
  
  # Distribution by borough
  filter(salary > 0 && salary < quantile(salary,0.75) + 1.5*IQR(salary)) %>%
  ggplot(aes(x=salary, color = BORO_NM)) + 
  geom_freqpoly() + ggtitle("Distribution of Annual Salaries \n by Borough")
```

## Covariation between variables
The two datasets need to be joined in order to analyze the relationship between income and crime.

```{r}
income_by_boro <- income_annual %>%
  mutate(BORO_NM=str_to_upper(boro)) %>%
  filter(BORO_NM == 'BRONX' | BORO_NM == 'BROOKLYN' | BORO_NM == 'QUEENS' |
         BORO_NM == 'MANHATTAN') %>%
  group_by(BORO_NM) %>%
  summarize(median_salary = median(salary))
```


```{r}
crime_by_boro <- complaints %>%
  select(CMPLNT_FR_DT, KY_CD, LAW_CAT_CD, BORO_NM) %>%
  filter(!is.na(BORO_NM)) %>%
  group_by(BORO_NM) %>%
  summarize(count = n())
```


```{r}
# income and crime
income_and_crime <- left_join(income_by_boro, crime_by_boro, by = "BORO_NM") %>%
  rename(complaint_count = count)
```


# Using geospatial data
```{r}
library(rgdal)
library(raster)
my_spdf <- shapefile("C:/Users/Student/OneDrive - University of Virginia/Desktop/SYS2202/Final/boro_shape_file/boro_shape_file.shp")
```
```{r}
library(broom)
# Tidying shape file
spdf_new <- tidy(my_spdf)

spdf_new <- spdf_new %>%
  mutate(id = recode(id, '0'='QUEENS', '1'='BROOKLYN', '2'='STATEN ISLAND', 
                     '3'='MANHATTAN', '4'='BRONX'))
```

```{r}
# Merge geospatial and numeric data
spdf_new <- spdf_new %>%
  left_join(. , crime_by_boro, by=c("id"="BORO_NM"))
```

```{r}
# Felonies
felonies <- complaints %>%
  filter(LAW_CAT_CD == "FELONY")
```


```{r}
# Plot map & points
library(ggplot2)
ggplot() +
  geom_polygon(data = spdf_new, aes(fill = count, x = long, y = lat, group = group)) +
  theme_void() +
  coord_map() +
  labs(fill = "Number of Crime Complaints") +
  geom_point(data=felonies, aes(x=Longitude, y=Latitude), size = 0.25, alpha = 0.75)
```



```{r}
spdf_new %>%
  left_join(. , complaints, by=c("long"="Longitude", "lat"="Latitude")) 
```

